### 2008-12-19 talk/ecoop2009/clibackend.tex
add a comment
minor fixes, some comments

Since in .NET methods are the basic units of compilation, a possible
solution consists in creating a new method 
any time a new case has to be added to a flexswitch.
\dacom{comment for Antonio: I am not sure this is the best solution. This cannot work for Java where classes are the basic
  units. Closures will be available only with Java Dolphin and I do
  not know how much efficient will be}
In this way, whereas flow graphs without flexswitches are translated
to a single method, the translation of flow graphs which can dynamically grow because of
flexswitches will be scattered over several methods.
Summarizing, the backend behaves in the following way:
\begin{itemize}
\item Each flow graph is translated in a collection of methods which
  can grow dynamically. \dacom{I propose primary/secondary instead of
    the overloaded terms main/child} Each collection contains at least one
  method, called \emph{primary}, which is the first to be created.
  All other methods, called \emph{secondary}, are added dynamically 
  whenever a new case is added to a flexswitch.

\item Each either primary or secondary method implements a certain
  number of blocks, all belonging to the same flow graph. Among these blocks
  there always exists an initial block whose input variables are
  parameters of the method; the input variables of all other blocks
  are local variables of the method.
\end{itemize}

When  a new case is added to a flexswitch, new blocks are generated
and translated by the backend in a new single method pointed
by a delegate which is stored in the code implementing the flexswitch,
so that the method can be invoked later.

@@

Following an internal link is not difficult in IL bytecode: a jump to
the corresponding code fragment in the same method is emitted 
to execute the new block, whereas the appropriate local variables are
used for passing arguments.
Also following an external link whose target is the initial block of a
method is not difficult: the corresponding method has to be invoked
with the appropriate arguments.

@@

To solve this problem every method contains a special code, called
\emph{dispatcher}; whenever a method is invoked, its dispatcher is
executed first\footnote{The dispatcher should not be
confused with the initial block of a method.} to
determine which block has to be executed.
This is done by passing to the method a 32 bits number, called 
\emph{block id}, which uniquely identifies the next block of the graph to be executed.
The high word of a block id is the id of the method to which the block
belongs, whereas the low word is a progressive number univocally identifying
each block implemented by the method.

@@

For instance, the code\footnote{For simplicity we write C\# code instead of
the actual IL bytecode.} generated for the dispatcher of method \texttt{0x0002}
is similar to the following fragment: 
\begin{small}
\begin{lstlisting}[language={[Sharp]C}]
// dispatch block
int methodid = (blockid && 0xFFFF0000) >> 16; 
int blocknum = blockid && 0x0000FFFF;         
if (methodid != MY_METHOD_ID) {
  // jump_to_ext 
  ...
}
switch(blocknum) {
  case 0: goto block0;
  case 1: goto block1;
  default: throw new Exception("Invalid block id");
}
\end{lstlisting}
\end{small}
If the next block to be executed is implemented in the same method
({\small\lstinline{methodid == MY_METHOD_ID}}), then the appropriate
jump to the corresponding code is executed, hence internal links
can be managed efficiently.
Otherwise, the \lstinline{jump_to_ext}
part of the dispatcher has to be executed.
The code that actually jumps to an external block is contained in
the dispatcher of the primary method, whereas the
\lstinline{jump_to_ext} code of dispatchers of secondary methods
simply delegates the dispatcher of the primary method of the same
graph (see later).

@@

Recall that the first argument of delegate \lstinline{FlexSwitchCase}
is the block id to jump to; since the target of an external jump is
always the initial block of the method, the first argument will be
always 0.

@@

\subsection{Alternative implementations}
\dacom{need to be discussed with Antonio}
\commentout{
Before implementing the solution described here, we carefully studied a lot of possible alternatives, but all of them either didn't work because of a limitation of the virtual machine or they could work but with terrible performances.

>>>>

Since in .NET methods are the basic units of compilation, a possible
solution consists in creating a new method 
any time a new case has to be added to a flexswitch.
In this way, whereas flow graphs without flexswitches are translated
to a single method, the translation of flow graphs which can dynamically grow because of
flexswitches will be scattered over several methods.
Summarizing, the backend behaves in the following way:
\begin{itemize}
\item Each flow graph is translated in a collection of methods which
  can grow dynamically. Each collection contains at least one
  method, called \emph{primary}, which is the first to be created.
  All other methods, called \emph{secondary}, are added dynamically 
  whenever a new case is added to a flexswitch.

\item Each either primary or secondary method implements a certain
  number of blocks, all belonging to the same flow graph. Among these blocks
  there always exists an initial block whose input variables are
  parameters of the method; the input variables of all other blocks
  are local variables of the method. \anto{This is wrong: the signature of the secondary methods is fixed, and input args are passed inside the InputArgs class, not as methodo parameters}
\end{itemize}

When  a new case is added to a flexswitch, new blocks are generated
and translated by the backend in a new single method pointed
by a delegate \footnote{\emph{Delegates} are the .NET equivalent of function pointers}
 of  which is stored in the code implementing the flexswitch,
so that the method can be invoked later.

@@

Following an internal link is not difficult in IL bytecode: a jump to
the corresponding code fragment in the same method is emitted 
to execute the new block, whereas the appropriate local variables are
used for passing arguments. \anto{this is wrong for the same reason as above}
Also following an external link whose target is the initial block of a
method is not difficult: the corresponding method has to be invoked
with the appropriate arguments.

@@

To solve this problem every method contains a special code, called
\emph{dispatcher}; whenever a method is invoked, its dispatcher is
executed first\footnote{The dispatcher should not be
confused with the initial block of a method.} to
determine which block has to be executed.
This is done by passing to the method a 32 bits number, called 
\emph{block id}, which uniquely identifies the next block of the graph to be executed.
The high word \anto{a word is 32 bit, block num and method id are 16 bit each} of a block id is the id of the method to which the block
belongs, whereas the low word is a progressive number univocally identifying
each block implemented by the method.

@@

For instance, the code\footnote{For simplicity we write C\# code instead of
the actual IL bytecode.} generated for the dispatcher of method \texttt{0x0002}
is similar to the following fragment: 
\begin{small}
\begin{lstlisting}[language={[Sharp]C}]
// dispatch block
int methodid = (blockid && 0xFFFF0000) >> 16; 
int blocknum = blockid && 0x0000FFFF;         
if (methodid != MY_METHOD_ID) {
  // jump_to_ext 
  ...
}
switch(blocknum) {
  case 0: goto block0;
  case 1: goto block1;
  default: throw new Exception("Invalid block id");
}
\end{lstlisting}
\end{small}
If the next block to be executed is implemented in the same method
({\small\lstinline{methodid == MY_METHOD_ID}}), then the appropriate
jump to the corresponding code is executed, hence internal links
can be managed efficiently. \anto{wrong: internal links don't go through the dispatcher}
Otherwise, the \lstinline{jump_to_ext}
part of the dispatcher has to be executed.
The code that actually jumps to an external block is contained in
the dispatcher of the primary method, whereas the
\lstinline{jump_to_ext} code of dispatchers of secondary methods
simply delegates the dispatcher of the primary method of the same
graph (see later).

@@

Recall that the first argument of delegate \lstinline{FlexSwitchCase} is the
block id to jump to. By construction, the target block of a flexswitch is
always the first in a secondary method, and we use the special value
\lstinline{0} to signal this.

@@

\subsection{Alternative implementations}
\dacom{need to be discussed with Antonio}

\anto{I propose to wait until the rest of paper is more or less in a final
  state, and see how much space is left}

\commentout{
Before implementing the solution described here, we carefully studied a lot of possible alternatives, but all of them either didn't work because of a limitation of the virtual machine or they could work but with terrible performances.

### 2008-12-19 talk/ecoop2009/jitgen.tex
minor fixes, some comments

\begin{figure}[h]
\label{fig:tlc-main}
\begin{center}
\input{tlc-simplified.py}
\caption{The main loop of the TLC interpreter}
\end{center}
\end{figure}

@@

The binding-time analyzer of our translation tool-chain is using a simple
abstract-interpretation based analysis. It is based on the
same type inference engine that is used on the source RPython program,
the annotator.  In this mode, it is called the \emph{hint-annotator}; it
operates over input graphs that are already low-level instead of
RPython-level, and propagates annotations that do not track types but
value dependencies and manually-provided binding time hints.
XXX the above needs rewriting when the background section is there

>>>>

\begin{figure}[h]
\label{fig:tlc-main}
\begin{center}
\input{tlc-simplified.py}
\caption{The main loop of the TLC interpreter, written in RPython}
\end{center}
\end{figure}

@@

The binding-time analyzer of our translation tool-chain is using a simple
abstract-interpretation based analysis. It is based on the
same type inference engine that is used on the source RPython program,
the annotator \anto{I'm not sure we should mention the annotator, as it is not referred anywhere else}.  In this mode, it is called the \emph{hint-annotator}; it
operates over input graphs that are already low-level instead of
RPython-level, and propagates annotations that do not track types but
value dependencies and manually-provided binding time hints.
XXX the above needs rewriting when the background section is there

### 2008-12-19 talk/ecoop2009/rainbow.tex
minor fixes, some comments

\section{Automatic Unboxing of Intermediate Results}

@@

Interpreters for dynamic languages typically continuously allocate a lot of small
objects, for example due to boxing. This makes arithmetic operations extremely
inefficient. For this reason, we
implemented a way for the compiler to try to avoid memory allocations in the
residual code as long as possible. The idea is to try to keep new
run-time structures "exploded": instead of a single run-time object allocated on
the heap, the object is "virtualized" as a set
of fresh variables, one per field. Only when the object can be accessed by from
somewhere else is it actually allocated on the heap. The effect of this is similar to that of
escape analysis \cite{XXX}, which also prevents allocations of objects that can
be proven to not escape a method or set of methods.

>>>>

\section{Automatic Unboxing of Intermediate Results}
\label{sec:virtuals}

@@

Interpreters for dynamic languages typically continuously allocate a lot of small
objects, for example due to boxing. This makes arithmetic operations extremely
inefficient. For this reason, we
implemented a way for the compiler to try to avoid memory allocations in the
residual code as long as possible. The idea is to try to keep new
run-time structures "exploded": instead of a single run-time object allocated on
the heap, the object is "virtualized" as a set
of fresh local variables, one per field. Only when the object can be accessed by from
somewhere else is it actually allocated on the heap. The effect of this is similar to that of
escape analysis \cite{XXX}, which also prevents allocations of objects that can
be proven to not escape a method or set of methods.
